Push-Pull Signal-Function Functional Reactive Programming

### 2012-11-27 Docs/IFL2012/introduction.tex
Break IFL paper into multiple files.

\documentclass[draft]{llncs}

\usepackage{fancyvrb}                                       % For code environment
\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small} %

\title{Push-Pull Signal-Function Functional Reactive Programming}
\titlerunning{Push-Pull Signal-Function FRP}

\author{Edward Amsden}
\institute{Rochester Institute of Technology\\\email{eca7215@cs.rit.edu}}

\begin{document}
\maketitle

\begin{abstract}
Functional Reactive Programming is a promising class of systems for writing
interactive and time-dependent programs. Signal-function FRP is a subclass of
these systems which provides advantages of modularity and correctness, but
has proven difficult to efficiently implement.

The abstraction of signal vectors provides the necessary type apparatus to
distinguish components of the input and output of signal functions which benefit
from a push-based implementation from those which benefit from a pull-based
implementation, and to combine both implementation strategies in a single system.

We describe a signal-function FRP system which provides push-based evaluation
for events, pull-based evaluation for signals, and a simple monadic evaluation
interface which permits the system to be easily integrated with one or more
IO systems.
\end{abstract}

@@

Section~\ref{section:System_Design} describes design choices for the system,
and provides an overview of the interface. Section~\ref{section:Implementation}
describes how the system is implemented, and how the separation of evaluation
between events and signals is achieved. Section~\ref{section:Discussion} is a
discussion of the usefulness of our implementation. 
Section~\ref{section:Ongoing_and_Further_Work} describes the current and future
work on this system. Section~\ref{section:Related_Work} gives an overview of
related efforts. Section~\ref{section:Conclusion} concludes.

\section{System Design and Interface}
\label{section:System_Design_and_Interface}

The goal of FRP is to provide an efficient, declarative abstraction for creating
reactive programs. Towards this overall goal, there are three goals which this
system is intended to meet.

Efficient evaluation is the motivation for push-based evaluation of events.
Since FRP programs are expected to  interact with an outside world in real time,
efficiency cannot simply be measured by the runtime of a program. Thus, when speaking of efficiency,
we are expressing a desire that the system utilize as few system resources as possible
for the task at hand, while responding as quickly as possible to external inputs and
producing output at a consistently high sample rate.

A composable abstraction is one in which values in that abstraction may be
combined in such a way that reasoning about their combined actions involves
little more than reasoning about their individual actions. In a signal function
system, the only interaction between composed signal functions ought to be that
the output of one is the input of another. Composability permits a particularly
attractive form of software engineering in which successively larger systems are
created from by combining smaller systems, without having to reason about the 
implementation of the components of the systems being combined.

It is fine for a system to be composable with regards to itself, but an FRP
system must interact with the outside world. Since we cannot anticipate every
possible form of input and output that the system will be asked to interact
with, we must interface with Haskell's IO system. In particular, most libraries
for user interaction (e.g. GUI and graphics libraries such as GTK+ and GLUT) and
most libraries for time-dependent IO (e.g. audio and video systems) make use of
the event loop abstraction. In this abstraction, event handlers are registered
with the system, and then a command is issued to run a loop which detects events
and runs the handlers, and uses the results of the handlers to render the
appropriate output.

We would like for the FRP system to be easy to integrate with such IO systems,
while being flexible enough to enable its use with other forms of IO systems,
such as simple imperative systems, threaded systems, or network servers.

\subsection{Semantics}
\label{subsection:System_Design_and_Interface-Semantics}

A rigorous and formal elucidation of the semantics of signal-function FRP remains
unattempted, but there is a sufficient change to the practical semantics of
signal-function FRP between previous signal-function systems and TimeFlies to warrant
some description.

In previous systems such as Yampa, events were understood (and implemented) as option-valued
signals. This approach is undesirable for several reasons. The most pressing reason is
that it prohibits push-based evaluation of events, because events are embedded in the
samples of a signal and must be searched for when sampling.

Another concern is that this approach limits the event rate to the sampling rate.
The rate of sampling should, at some level, not matter to the FRP system. Events
which occur between sampling intervals are never observed by the system. Since
the signal function is never evaluated except at the sampling intervals, no
events can otherwise be observed.

This concern drives the next concern. Events are not instantaneous in this formulation.
If a signal is option valued, the sampling instant must fall within the interval where
there is an event occurrence present for that event to be observed. If events are
instantaneous, the probability of observing an event occurrence is zero.

Therefore, TimeFlies employs the N-Ary FRP type formulation to represent signals and
events as distinct entities in the inputs and outputs of signal functions. This means
we are now free to choose our representation of events, and to separate it from the
representation and evaluation of signals.

This freedom yields the additional ability to make events independent of the sampling
interval altogether. The semantics of event handling in TimeFlies is that an event occurrence
is responded to immediately, and does not wait for the next sampling instant. This allows events
to be instantaneous, and further, allows multiple events to occur within a single sampling interval.

There are two tradeoffs intrinsic to this approach. The first is that events are only partially ordered
temporally. There is no way to guarantee the order of observation of event occurrences occurring in the
same sampling interval. Further, the precise time of an event occurrence cannot be observed, only the 
time of the last sample prior to the occurrence.

In return for giving up total ordering and precise observation of the timing of events, we obtain the
ability to employ push-based evaluation for event occurrences, and the ability to non-deterministically
merge event occurrences. When events being input to a non-deterministic merge have simultaneous occurrences,
we arbitrarily select one to occur first. This does not violate any guarantee about time values, since
they will both have the same time value in either case, and does not violate any guarantee about ordering,
since no guarantee of their order is given.

A formal semantic description of signal function FRP would clarify the consequences of this decision somewhat,
but is outside the scope of this thesis.

\subsection{Types}
\label{subsection:System_Design_and_Interface-Types}

In a strongly and statically typed functional language, types are a key part of
an interface. Types provide a mechanism for describing and ensuring properties
of the interface's components and about the systems created with these
components.

In order to type signal functions, we must be able to describe their input and
output. In most signal function systems, a signal function takes exactly one
input and produces exactly one output. Multiple inputs or outputs are handled
by making the input or output a tuple, and combinators which combine or split
the inputs or outputs of a signal assume this. Events are represented at the
type level as a particular type of signal, and at the value level as an option,
either an event occurrence or not.

This method of typing excludes push-based evaluation at the outset.
It is not possible to construct a ``partial tuple'' nor in general is it
possible to construct only part of any type of value. Push-based evaluation
depends on evaluating only that part of the system which is updated, which means
evaluating only that part of the input which is updated.

In order to permit the construction of partial inputs and outputs, we make use
of signal vectors. Signal vectors are uninhabited types which describe the input
and output of a signal function. Singleton vectors are parameterized over the
type carried by the signal or by event occurrences. The definition of the signal
vector type is shown in Figure~\ref{figure:signal_vector_types}.

Having an uninhabited signal vector type allows us to construct representations
of inputs and outputs which are hidden from the user of the system, and are
designed for partial representations.

\begin{figure}
\begin{code}
data SVEmpty    -- An empty signal vector component,
                -- neither event nor signal
data SVSignal a -- A signal, carrying values of type a
data SVEvent a  -- An event, whose occurrences carry values of type a
data SVAppend svLeft svRight -- The combination of the signal vectors
                             -- svLeft and svRight
\end{code}
\hrule
\caption{Signal vector types.}
\label{figure:signal_vector_types}
\end{figure}

The type constructor for signal functions is shown in
Figure~\ref{figure:signal_function_types}. For the {\tt init} parameter, only
one possible instantiation is shown. The usefulness of this type parameter,
along with another instantation which is hidden from users of the library,
is discussed in the section on implementation of signal functions
(Section~\ref{section:Implementation}).

Signal functions with signal vectors as input and output types form a
Haskell {\tt GArrow}~\cite{Megacz2011}. Specifically, the signal function
type constructor (with the initialization parameter fixed) forms the arrow
type, the {\tt SVAppend} type constructor forms the product type, and the
{\tt SVEmpty} type constructor forms the unit type.

The representation of signal functions is discussed in
Section~\ref{section:Implementation-Signal_Functions}. The type synonyms
{\tt :\textasciitilde>} and {\tt :\textasciicircum:} are included for
readability and are not crucial to the FRP system.

\begin{figure}
\begin{code}
-- Signal functions
-- init: The initialization type for 
--   the signal function, always NonInitialized
--   for exported signal functions
-- svIn: The input signal vector
-- svOut: The output signal vector
data SF init svIn svOut

data NonInitialized

type svIn :~> svOut = SF NonInitialized svIn svOut
type svLeft :^: svRight = SVAppend svLeft svRight
\end{code}
\hrule
\caption{Signal function types.}
\label{figure:signal_function_types}
\end{figure}

Monad transformers~\cite{Jones1995} provide a means to combine the functionality
of multiple monads. A monad transformer is a monad with an extra type parameter.
This type parameter is instantiated with the type constructor of the underlying
monad, and an extra operation {\tt lift} is provided which converts values in
the underlying monad to values in the monad transformer.

The evaluation monad is a monad transformer. This permits it to be used in
conjunction with Haskell's {\tt IO} monad (or any other monad) to describe how
input is to be obtained for the signal function being evaluated, and how outputs
are to be handled.

The evaluation monad, in addition to the standard monad operators, provides a
means of {\em initializing} a signal function, and a means of translating the
monadic value describing evaluation to a value in the underlying monad. This
means, for instance, that we can obtain an action in the {\tt IO} monad to
evaluate a signal function.

The type of the evaluation monad must track the input and output type of the
signal function. The monad's context stores a mapping from outputs to handling
actions. There are thus four type parameters to the monad's type
constructor: the input signal vector, the output signal vector, the type of the
underlying monad, and the monadic type parameter. The type is shown in
Figure~\ref{figure:evaluation_monad_types}.

\begin{figure}
\begin{code}
-- A signal function's evaluations state
data SFEvalState svIn svOut m
-- The evaluation monad
data SFEvalT svIn svOut m a

instance Monad m => Monad (SFEvalT svIn svOut m)
\end{code}
\hrule
\caption{Evaluation monad types.}
\label{figure:evaluation_monad_types}
\end{figure}

\subsection{Combinators}
\label{section:System_Design_and_Interface-Combinators}

Signal functions are constructed from combinators, which are primitive signal
functions and operations to combine these primitives. These combinators are
grouped as basic signal functions, lifting operations for pure functions,
routing, reactivity, feedback, event processing, joining, and time dependence.

The basic signal functions (Figure~\ref{figure:basic_signal_functions})
provide very simple operations. The {\tt identity} signal function, as expected,
simply copies its input to its output. The {\tt constant} signal function
produces the provided value as a signal at all times. The {\tt never} signal
function has an event output which never produces occurrences. The {\tt asap}
function produces an event occurrence with the given value at the first time
step after it is switched into the network. The {\tt after} function waits for
the specified amount of time before producing an event occurrence.

With the exception of {\tt identity}, all of the basic signal functions have
empty inputs. This allows these signal functions to be used to insert values
into the network which are known when the signal function is created, without
having to route those values from an input.

\begin{figure}
\begin{code}
-- Pass the input unmodified to the output
identity :: sv :~> sv

-- Produce a signal which is at all times the supplied value
constant :: a -> SVEmpty :~> SVSignal a

-- An event with no occurrences
never    :: SVEmpty :~> SVEvent a

-- An event with one occurrence, as soon as possible after
-- the signal function is initialized
asap     :: a -> SVEmpty :~> SVEvent a

-- An event with one occurrence
-- after the specified amount of time has elapsed.
after    :: Double -> a -> SVEmpty :~> SVEvent a
\end{code}
\hrule
\caption{Basic signal function combinators.}
\label{figure:basic_signal_functions}
\end{figure}

Two combinators are provided to lift pure functions to signal functions
(Figure~\ref{figure:lifting_pure_functions}). The {\tt pureSignalTransformer}
combinator applies the pure function to a signal at every sample point. The
{\tt pureEventTransformer} combinator applies the function to each occurrence of
an input event.

\begin{figure}
\begin{code}
-- Apply the given function to a signal at all points in time
pureSignalTransformer :: (a -> b) -> SVSignal a :~> SVSignal b

-- Apply the given function to each event occurrence
pureEventTransformer  :: (a -> b) -> SVEvent a :~> SVEvent b
\end{code}
\hrule
\caption{Lifting pure functions.}
\label{figure:lifting_pure_functions}
\end{figure}

The routing combinators are used to combine signal functions, and
to re-arrange signal vectors in order to connect signal functions.
The routing combinators are shown in Figure~\ref{figure:routing_combinators}.

Only those combinators which modify or combine signal functions
({\tt (>>>)}, {\tt first}, {\tt second}) are reactive (may replace themselves in
response to events, and then only if they inherit their reactivity from the
signal function(s) which are inputs to combinator. The rest do not
react to or modify the input in any way, except to re-arrange it, copy it, or
discard it altogether.

\begin{figure}
\begin{code}
-- Use the output of one signal function as the input for another
(>>>) :: (svIn :~> svBetween) -> (svBetween :~> svOut) -> svIn :~> svOut

-- Pass through the right side of the input unchanged
first :: (svIn :~> svOut) -> (svIn :^: sv) :~> (svOut :^: sv)

-- Swap the left and right sides
swap :: (svLeft :^: svRight) :~> (svRight :^: svLeft)

-- Duplicate the input
copy :: sv :~> (sv :^: sv)

-- Ignore the input
ignore :: sv :~> svEmpty

-- Remove an empty vector on the left
cancelLeft :: (SVEmpty :^: sv) :~> sv

-- Add an empty vector on the left
uncancelLeft :: sv :~> (SVEmpty :^: sv)

-- Make right-associative
associate :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3))
\end{code}
\hrule
\caption{Routing combinators.}
\label{figure:routing_combinators}
\end{figure}

Reactivity is introduced by means of the {\tt switch} combinator
(Figure~\ref{figure:switch_combinator}). The design of this combinator
allows modular packaging of reactivity. A signal function can determine
autonomously when to replace itself, based only on its input and state,
by emitting an event occurrence carrying its replacement. The combinator
consumes and hides the event carrying the replacement signal function, 
so the reactivity is not exposed by the resulting reactive signal function.

\begin{figure}
\begin{code}
switch ::    (svIn :~> (svOut :^: SVEvent (svIn :~> svOut)))
          -> svIn :~> svOut
\end{code}
\hrule
\caption{Reactivity combinator.}
\label{figure:switch_combinator}
\end{figure}

It is often useful for a signal function to receive a portion of its
own output as input. This is especially useful when we have two
signal functions which we would like to mutually interact. We cannot
just serially compose them, we must also bring the output of the second
back around to the first. Many signal-processing algorithms also depend
on feedback. The combinator which provides this ability is shown
in Figure~\ref{figure:feedback_combinator}.

\begin{figure}
\begin{code}
loop ::    ((svIn :^: svLoop) :~> (svOut :^: svLoop))
        -> svIn :~> svOut
\end{code}
\hrule
\caption{Feedback combinator.}
\label{figure:feedback_combinator}
\end{figure}

This combinator provides decoupling for signals
(the input signal is the output signal at the previous time-step)
but not events (event occurrences are supplied to the combinator immediately).
This means that the programmer has the responsibility to ensure that feedback
does not generate an infinite sequence of events in a single time-step.

Several combinators are provided for manipulating, suppressing, and generating events.
Each of the combinators has an option variant and a list variant. The option variant
produces an output event occurrence whenever the application of the supplied function
to the input event produces a value. The list version produces an event occurrence for
each of the elements of the output list. The combinators are shown in Figure~\ref{figure:event_specific_combinators}.

\begin{figure}
\begin{code}
-- Apply the function to each input occurrence,
-- and produce an occurrence for each Just.
filter :: (a -> Maybe b) -> SVEvent a :~> SVEvent b

-- Apply the function to the stored accumulator
-- and the event occurrence, replacing the accumulator
-- and possibly outputting an occurrence
accumulate ::    (a -> b -> (Maybe c, a))
              -> a
              -> SVEvent b :~> SVEvent c
\end{code}
\hrule
\caption{Event-specific combinators.}
\label{figure:event_specific_combinators}
\end{figure}

The filter combinators are stateless, and thus apply the function to only the new
input value. They are useful for suppressing events, as well as for extracting one
of multiple cases of a datatype. For instance, a splitter for events carrying
{\tt Either}-valued occurrences could be written as:

\begin{code}
getLeft :: Either a b -> Maybe a
getLeft (Left x) = Just x
getLeft _ = Nothing

getRight :: Either a b -> Maybe b
getRight (Right x) = Just x
getRight _ = Nothing

split :: SVEvent (Either a b) :~> (SVEvent a :^: SVEvent b)
split = copy >>> first (filter getLeft) >>> second (filter getRight)
\end{code}

The accumulate combinators are stateful, applying the supplied function
to both the input value and an accumulator. This function has two results:
the option or list of output event occurrence values, and the new value
for the accumulator.

The accumulator is useful when responses to multiple event occurrences
(from one or more sources) must be coordinated. For instance, in the
benchmark application (see Chapter~\ref{chapter:Evaluation_and_Comparisons}),
a table is maintained that allows knowledge from previous event occurrences
(packets from a network switch) to be used in deciding where the present
packet ought to go.

The joining combinators provide the ability to combine two event
streams, two signals, or a signal and an event stream. These
combinators are shown in Figure~\ref{figure:joining_combinators}

The {\tt union} combinator is a non-deterministic merge of event
streams. Any event which occurs on either input will occur
on the output. For simultaneous event occurrences, the order of occurrence
is not guaranteed, but the occurrence itself is. This construct
is also guaranteed to respect the relationship of event occurrences to sampling
intervals.

The {\tt combineSignals} combinator applies a binary function pointwise to two
signals, and produces the result of this application as a third signal. The
combining function is necessary because we will have two input samples at each
time step, and must produce exactly one output sample.

The {\tt capture} combinator adds the last-sampled value of a signal at the time
of an event occurrence to that event occurrence.

These three combinators together provide the ability to combine elements of
signal vectors. By combining these combinators, along with the {\tt cancelLeft}
and {\tt cancelRight} routing combinators, arbitrary signal vectors can be
reduced.

\begin{figure}
\begin{code}
union          :: (SVEvent a :^: SVEvent a) :~> SVEvent a
combineSignals :: (a -> b -> c) -> (SVSignal a :^: SVSignal b) :~> SVSignal c
capture        :: (SVSignal a :^: SVEvent b) :~> SVEvent (b, a)
\end{code}
\hrule
\caption{Joining combinators.}
\label{figure:joining_combinators}
\end{figure}

A set of combinators are provided for making use of time-dependence in a signal
function. These combinators allow the modification of signals and events with
respect to time, and the observation of the current time value. The combinators
are shown in Figure~\ref{figure:time_combinators}

\begin{figure}
\begin{code}
time      :: SVEmpty :~> SVSignal Double
delay     :: Double -> (SVEvent a :^: SVEvent Double) :~> SVEvent a
integrate :: TimeIntegrate i => SVSignal i :~> SVSignal i
\end{code}
\hrule
\caption{Time-dependent combinators.}
\label{figure:time_combinators}
\end{figure}

The simplest time-dependent combinator is {\tt time}, which simply outputs
the time since it began to be evaluated. This does not necessarily correspond to
the time since the global signal function began to be evaluated, since the
signal function in which the {\tt time} combinator is used may have been
introduced through {\tt switch}.

The {\tt delay} signal function allows events to be delayed. An initial delay
time is given, and event occurrences on the right input can carry a new delay
time. Event occurrences on the left input are stored and output when their delay
time has passed. Changing the delay time does not affect occurrences already
waiting.

The {\tt integrate} combinator outputs the rectangle-rule approximation of the
integral of its input signal with respect to time.

\subsection{Evaluation Interface}
\label{subsection:System_Design_and_Interface-Evaluation_Interface}

The evaluation interface provides a modified state monad which holds a signal
function, together with some additional information, as its state (shown in Figure~\ref{figure:evaluation_state}.
Rather than monadic instructions to put and get the state, the monad provides instructions
to trigger an input event, update an input signal, and trigger sampling of
signals in the signal function. Additional state includes the current set of
modifications to the input signals (since the last sample) and a set of
handlers which actuate effects based on output events or changes to the output
signal.

\begin{figure}
\begin{code}
-- A vector of handlers for outputs
data SVHandler out sv

-- The state maintained when evaluating a signal function
data SFEvalState m svIn svOut

-- Create the initial state for evaluating a signal function
initSFEval ::    SVHandler (m ()) svOut   -- Output handlers

              -> SVSample svIn            -- Initial input samples

              -> Double                   -- Initial external time,
                                          -- corresponding to time 0 for
                                          -- the signal function

              -> (svIn :~> svOut)         -- Signal function to evaluate
              -> SFEvalState m svIn svOut
\end{code}
\hrule
\caption{State maintained when evaluating a signal function.}
\label{figure:evaluation_state}
\end{figure}

Sets which correspond to signal vectors are built with ``smart'' constructors.
For instance, to construct a set of handlers, individual handling functions are
lifted to handler sets with the {\tt signalHandler} and {\tt eventHandler}
functions, and then combined with each other and {\tt emptyHandler} leaves
using the {\tt combineHandlers} function.

Building the initial input sample is similar, but {\tt sampleEvt} leaves do
not carry a value.

In order to initialize the state, the user must supply a set of handlers, the
signal function to evaluate, and initial values for all of the signal inputs
(Figure~\ref{figure:initial_input}).

This state can then be passed to a monadic action which will supply input to
the signal function. Inputs are constructed using a simple interface with
functions to construct sample updates and event occurrences, and to specify
their place in the vector (Figure~\ref{figure:ongoing_input}).

\begin{figure}
\begin{code}
-- Class to overload left and right functions
class SVRoutable r where
  svLeft          :: r svLeft -> r (svLeft :^: svRight)
  svRight         :: r svRight -> r (svLeft :^: svRight)

-- An input event occurrence
data SVEventInput sv
instance SVRoutable SVEventInput sv

-- An updated sample for a signal
data SVSignalUpdate sv
instance SVRoutable SVSignalUpdate sv

-- Create an event occurrence
svOcc           :: a -> SVEventInput (SVEvent a)

-- Create an updated sample
svSig           :: a -> SVSignalUpdate (SVSignal a)
\end{code}
\hrule
\caption{Data types for ongoing input.}
\label{figure:ongoing_input}
\end{figure}

The {\tt SFEvalT} monad is actually a monad transformer, that is, it is
parameterized over an underlying monad whose actions may be lifted to
{\tt SFEvalT}. In the usual case, this will be the {\tt IO} monad.

{\tt SFEvalT} actions are constructed using combinators to push events,
update inputs, and step time, as well as actions lifted from the underlying
monad (used to obtain these inputs). An action in the underlying monad
which produces the result and a new state is obtained with the {\tt runSFEvalT}
function. These combinators are shown in Figure~\ref{figure:evaluation_combinators}.

\begin{figure}
\begin{code}
-- The evaluation monad
data SFEvalT svIn svOut m a
-- Appropriate instances for Monad, MonadTrans, MonadIO,
-- Functor, Applicative, etc

-- Obtain an action in the underlying monad
-- from an SFEvalT and a new state.
runSFEvalT ::    SFEvalT svIn svOut m a
              -> SFEvalState m svIn svOut
              -> m (a, SFEvalState m svIn svOut)

-- Push an event occurrence.
push :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()

-- Update the value of an input signal sample
-- (not immediately observed)
update :: (Monad m) => SVEventInput svIn -> SFEvalT svIn svOut m ()

-- Step forward in time, observing the updated signal values
step :: (Monad m) => Double -> SFEvalT svIn svOut m ()
\end{code}
\hrule
\caption{Evaluation combinators}
\label{figure:evaluation_combinators}
\end{figure}

\section{Implementation}
\label{section:Implementation}

We now turn our attention to the implementation of the signal function
system. We will discuss representations of inputs, outputs, and signal functions,
as well as the implementations of specific signal function combinators.

\subsection{Signal Functions}
\label{subsection:Implementation-Signal_Functions}

The design of signal functions specifies a family of types for the inputs and
outputs of signal functions. Signal functions are not functions in the purest
sense, however. They are not mappings from a single instance of their input
type to a single instance of their output type. They must be implemented with
respect to the temporal semantics of their inputs and outputs.

We therefore start by creating a set of concrete datatypes for the inputs and
outputs of signal functions. These datatypes will be parameterized by the input
and output types of the signal function, and will not be exposed to the user of
the library. Rather, they will specify how data is represented during the
temporal evaluation of signal functions. We then implement signal functions
as records of functions from these concrete types to these concrete types paired
with new signal functions. The evaluation system for signal functions then
maintains this record for the signal function it is evaluating, calls the
correct function for the current input, actuates the outputs (using supplied
functions from the outputs to monadic actions) and replaces the record with the
newly supplied signal function record.

In Section~\ref{section:System_Design_and_Interface} we presented signal vectors
as a set of types. In order to be completely principled, we should isolate these
types into their own {\em kind} (a sort of type of types); however, the Haskell
extension for this was far from stable at the time this system was created.

We create concrete representations which are parameterized over
these types. These concrete representations will be expressed as GADTs, allowing
each data constructor of the representation to fill in a specific signal vector
type for the parameter of the representation.

The first thing to represent is {\em samples}, which are sets of values for
the signal components of a signal vector. Therefore, we create a representation
which carries a value for every {\tt SVSignal} leaf of a signal vector. In order
to do this, we restrict each of our constructors to a single signal vector type.
So there are three leaf constructors: {\tt SVSample}, which carries a value; and
{\tt SVSampleEvent} and {\tt SVSampleEmpty}, which do not. This ensures that the
only way to represent a sample leaf is with the {\tt SVSample} constructor,
which carries a value of the appropriate type. The datatype is shown in
Figure~\ref{figure:signal_sample_datatype}.

\begin{figure}
\begin{code}
data SVSample sv where
  SVSample      ::    a
                   -> SVSample (SVSignal a)
  SVSampleEvent ::    SVSample (SVEvent a)
  SVSampleEmpty ::    SVSample SVEmpty
  SVSampleBoth  ::    SVSample svLeft
                   -> SVSample svRight
                   -> SVSample (SVAppend svLeft svRight)
\end{code}
\hrule
\caption{Datatype for signal samples.}
\label{figure:signal_sample_datatype}
\end{figure}

What about the event components? We want to represent event occurrences,
each of which will correspond to at most one event in the vector. So a different
representation is called for. In this case, there will be only three
constructors. One constructor will represent an event leaf, and the other will
represent a single value on the left or right side of the node ({\tt SVAppend}),
ignoring all of the type structure on the other side. This representation
describes a path from the root of the signal vector, terminating at an event
leaf with a value.

By pattern matching on the path constructors, we can determine which subvector
of a signal vector an event occurrence belongs to, repeatedly refining it until
we determine which event in the vector the occurrence corresponds to. The
datatype for occurrences is shown in Figure~\ref{figure:event_occurrence_datatype}.

\begin{figure}
\begin{code}
data SVOccurrence sv where
  SVOccurrence ::    a
                  -> SVOccurrence (SVEvent a)
  SVOccLeft    ::    SVOccurrence svLeft
                  -> SVOccurrence (SVAppend svLeft svRight)
  SVOccRight   ::    SVOccurrence svRight 
                  -> SVOccurrence (SVAppend svLeft svRight)
\end{code}
\hrule
\caption{Datatype for event occurrences.}
\label{figure:event_occurrence_datatype}
\end{figure}

We add one more representation for signals, in order to avoid uneccessary
representations of the values of all signals when not all signals have changed
their values. This representation allows us to represent the values of zero or
more of the signals in a signal vector. To accomplish this, we replace the
individual constructors for the {\tt SVEmpty} and {\tt SVEvent} leaves with %there is a better word, something about "useless but not really", for this
a single, unconstrained constructor. This constructor can represent an arbitrary
signal vector. We can use the constructor for signal vector nodes and the 
constructor for sample leaves to represent the updated values, while filling
in the unchanged portions of the signal vector with this general constructor.
This datatype is shown in Figure~\ref{figure:signal_update_datatype}.

\begin{figure}
\begin{code}
data SVDelta sv where
  SVDeltaSignal  ::    a
                    -> SVDelta (SVSignal a)
  SVDeltaNothing ::    SVDelta sv
  SVDeltaBoth    ::    SVDelta svLeft
                    -> SVDelta svRight
                    -> SVDelta (SVAppend svLeft svRight)
\end{code}
\hrule
\caption{Datatype for signal updates.}
\label{figure:signal_update_datatype}
\end{figure}

We now have concrete datatypes for an implementation to operate on. Our next
task is to represent transformers of temporal data, which themselves may change
with time. The common approach to this task is sampling, in which a program
repeatedly checks for updated information, evaluates it, updates some state,
and produces an output. This is the essence of pull-based evaluation.

Another approach is notification, in which the program exposes an interface
which the source of updated information may invoke. This is a repeated entry
point to the program, which causes the program to perform the same tasks
listed above, namely, evaluate the updated information, update state, and
produce output. The strategy of notification as opposed to repeated checking is
the essence of push-based evaluation.

Signal functions are declarative objects, and not running processes. They have
no way to invoke sampling themselves. They can, however, expose separate
interfaces for when sampling is invoked, and when they are notified of an event
occurrence. This creates two control paths through a signal function. One of
these control paths is intended to be invoked regularly and frequently with
updates to the time and sample values, and the other is intended to be invoked
only when an event occurs. The benefit of separating these control paths is that
events are no longer defined in terms of sampling intervals, and need not even
be considered in sampling, except when they are generated by a condition on a
sample. On the other hand, events can be responded to even if the time has not
yet come for another sample, and multiple events can be responded to in a single
sampling interval.

We represent signal functions as a GADT with three type parameters and two 
constructors. The first type parameter represents the initialization state,
and is specialized to {\tt Initialized} or {\tt NonInitialized} depending on the
constructor. The other two type parameters are the input and output signal
vectors, respectively. The signal functions that a user will compose are\
non-initialized signal functions. They must be provided with an initial set of
input signals (corresponding to time zero). When provided with this input, they
produce their time-zero output, and an initialized signal function. The datatype
is shown in Figure~\ref{figure:signal_function_datatype}.

\begin{figure}
\begin{code}
data Initialized

data NonInitialized

data SF init svIn svOut where
  SF     ::    (SVSample svIn 
                  -> (SVSample svOut,
                      SF Initialized svIn svOut)) 
            -> SF NonInitialized svIn svOut
  SFInit ::    (Double 
                  -> SVDelta svIn
                  -> (SVDelta svOut,
                      [SVOccurrence svOut],
                      SF Initialized svIn svOut)) 
            -> (SVOccurrence svIn
                  -> ([SVOccurrence svOut],
                      SF Initialized svIn svOut))
            -> SF Initialized svIn svOut
\end{code}
\hrule
\caption{Datatype and empty types for signal functions.}
\label{figure:signal_function_datatype}
\end{figure}

Initialized signal functions carry two continuations. The first continuation
takes a time differential and a set of signal updates, and returns a set of
signal updates, a collection of event occurrences, and a new initialized signal
function of the same type. This is the continuation called when sampling.

The second continuation takes an event occurrence, and returns a collection of
event occurrences and a new signal function of the same type. This continuation
is only called when there is an event occurrence to be input to the signal
function.

Note that each of these continuations uses one or more of the concrete
representations of signal vectors, and applies the type constructor for the
representation to the input or output signal vector for the signal function.

Having specified a type constructor for signal functions, we must now provide
combinators which produce signal functions typed with this type constructor.
Space precludes a complete description of the combinators in the TimeFlies
system, but we will present

The simplest example of the implementation of is the {\tt identity} combinator,
shown in Figure~\ref{figure:identity_implementation}. This signal function
simply passes all of its inputs along as outputs. The initialization function
simply passes along the received sample and outputs the initialized version of
the signal function. The initialized version of the input is similar, but is
self-referential. It outputs itself as its replacement. This is standard for
simple and routing combinators which are not reactive, and simply re-arrange,
discard, or combine signals and events.

In order for our primitive signal functions to be useful, we need a means of
composing them. Serial composition creates one signal function from two, by
using the output of one as the input of the other. The serial composition
combinator is styled {\tt (>>>)}. The implementation of this operator is one
place where the advantage of responding to events independently from signal
samples becomes clear.

This is the only primitive combinator which takes two signal functions, and
thus, it is the only way to combine signal functions. Parallel, branching, and
joining composition can be achieved by modifying signal functions with the
{\tt first} and {\tt second} combinators and composing them with the
routing and joining combinators.

Combinators which take one or more signal functions as input must recursively
apply themselves, as is shown in the implementation of serial composition
(Figure~\ref{figure:serial_composition_implementation}). They must also
handle initialization, retaining the initialized signal functions and passing
them to the initialized version of the combinator.

The switch combinator is the means of introducing reactivity into a signal
function. This combinator allows a signal function to replace itself by
producing an event occurrence. The combinator wraps a signal function, and 
observes an event on the right side of the output signal vector. At the first
occurrence of the event, the signal function carried by the occurrence replaces
the signal function.

The switch combinator stores the input sample provided during initialization,
and updates it with the input signal updates. When the wrapped signal function
produces an occurrence carrying a new signal function, that signal function is
initialized with the stored input sample. It is then "wrapped" by another
function which closes over its output sample, and outputs the sample as a signal
update as the next time step. After this, it acts as the new signal function.
This wrapping has some performance implications, which are discussed

This combinator checks the outputs of the wrapped
signal function for an event occurrence from which an uninitialized signal
function is extracted. The switch combinator stores the full sample
for its input vector (which is identical to the input vector of the supplied
signal function) to initialize the new signal function. This also demands that
it add a wrapper to the new signal function which waits for the next sampling
interval and actuates the sample output at initialization as an output set
of changes to the signal. This has some performance implications, which are
discussed in Chapter~\ref{chapter:Evaluation_and_Comparisons}.

Most of the routing combinators are simple to implement. The only task is to add
remove, or replace routing constructors on signal updates and event occurrences.
Since these signal functions are stateless and primitive, they can simply
return themselves as their replacements. The {\tt swap} combinator is shown as
an example in Figure~\ref{figure:swap_implementation}

The {\tt first} and {\tt second} combinators are similar to serial composition,
but they transform only one signal function. For signal changes, they must split
the set of input changes into those which will be passed to the signal function
and those which will be simply passed along to the output, and then recombine
them on the other side. For event occurrences, the occurrence must be
pattern-matched to determine whether to call the event continuation from the
provided signal function or passed through, and output event occurrences must
have the suitable routing constructor re-applied. In any case, when a
continuation has been applied, the combinator must be recursively applied to the
new signal function.

The looping feedback combinator is particularly tricky. As it is currently
implemented, the initial sample for the right side of the input signal vector to
the supplied function is the right side of the output sample. This is acceptable,
given Haskell's non-strict evaluation strategy, but it is necessary that the
right side of the signal function's output not be immediately dependent on its
input. The feedback combinator makes use of Haskell's lazy evaluation to
feed events back into the combinator, and stores signal updates until the next
sample. Signal samples are thus automatically decoupled after initialization.
The implementation makes use of the recursive nature of the {\tt let} construct
in Haskell, and the non-strict evaluation of Haskell, to implement feedback.

The {\tt filter} combinators are simple to implement. Their sampling
continuation is superfluous, and the event continuation merely applies the
supplied function, and constructs an output list based on the result.

The {\tt accumulate} combinators are implemented in terms of the {\tt filter}
and {\tt switch} combinators, as shown in
Figure~\ref{figure:accumulate_implementation}.

>>>>

@@

Section~\ref{section:System_Design_and_Interface} describes design choices for the system,
and provides an overview of the interface. Section~\ref{section:Implementation}
describes how the system is implemented, and how the separation of evaluation
between events and signals is achieved. Section~\ref{section:Discussion} is a
discussion of the usefulness of our implementation. 
Section~\ref{section:Ongoing_and_Further_Work} describes the current and future
work on this system. Section~\ref{section:Related_Work} gives an overview of
related efforts. Section~\ref{section:Conclusion} concludes.
